---
title: "Relationship between DMS data and sarbecovirus RBD natural diversity"
author: "Tyler Starr"
date: "5/18/2020"
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: inline
---

This notebook analyzes DMS meausurements in relation to natural sequence diversity in a broader alignment of sarbecovirus RBDs.

## Setup

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

#list of packages to install/load
packages = c("yaml","data.table","tidyverse","gridExtra","bio3d")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages])
}
#load packages
invisible(lapply(packages, library, character.only=T))

#read in config file
config <- read_yaml("config.yaml")

#read in file giving concordance between RBD numbering and SARS-CoV-2 Spike numbering
RBD_sites <- data.table(read.csv(file="data/RBD_sites.csv",stringsAsFactors=F))

#make output directory
if(!file.exists(config$sarbecovirus_diversity_dir)){
  dir.create(file.path(config$sarbecovirus_diversity_dir))
}
```

Session info for reproducing environment:
```{r print_sessionInfo}
sessionInfo()
```

Read in tables of variant effects on binding and expression for single mutations to the SARS-CoV-2 RBD and for homolog RBDs.

```{r read_data}
homolog_measurements <- data.table(read.csv(file=config$homolog_effects_file,stringsAsFactors = F))
mutants <- data.table(read.csv(file=config$single_mut_effects_file,stringsAsFactors = F))

#rename mutants site indices to prevent shared names with RBD_sites, simplifying some downstream calculations that cross-index these tables
setnames(mutants, "site_RBD", "RBD_site");setnames(mutants, "site_SARS2", "SARS2_site")

#add color column to homologs, by clade
homolog_measurements$clade_color <- as.character(NA); homolog_measurements[clade=="Clade 1",clade_color := "#EF4136"]; homolog_measurements[clade=="Clade 2",clade_color := "#009444"]; homolog_measurements[clade=="Clade 3",clade_color := "#EE2A7B"]; homolog_measurements[clade=="SARS-CoV-2",clade_color := "#2E3192"]

#add mean, max, min mut effects per site annotations
RBD_sites[,mean_bind := mean(mutants[SARS2_site==site_SARS2 & wildtype != mutant & mutant != "*",bind_avg],na.rm=T),by=site_SARS2]
RBD_sites[,max_bind := max(mutants[SARS2_site==site_SARS2 & wildtype != mutant & mutant != "*",bind_avg],na.rm=T),by=site_SARS2]
RBD_sites[,min_bind := min(mutants[SARS2_site==site_SARS2 & wildtype != mutant & mutant != "*",bind_avg],na.rm=T),by=site_SARS2]

RBD_sites[,mean_expr := mean(mutants[SARS2_site==site_SARS2 & wildtype != mutant & mutant != "*",expr_avg],na.rm=T),by=site_SARS2]
RBD_sites[,max_expr := max(mutants[SARS2_site==site_SARS2 & wildtype != mutant & mutant != "*",expr_avg],na.rm=T),by=site_SARS2]
RBD_sites[,min_expr := min(mutants[SARS2_site==site_SARS2 & wildtype != mutant & mutant != "*",expr_avg],na.rm=T),by=site_SARS2]
```

## Relationship between DMS mutational tolerance and natural sequence diversity

We load in an alignment of sarbecovirus RBD sequences, whose curation and alignment is described in the `data/alignments` subdirectory. Many of these sequences are so-called "Clade 2" sequences, which have not been shown to bind human or any other ACE2 -- so whether they evolve under constraint for ACE2-binding is unclear. Therefore, we consider the entire sarbecovirus RBD alignment, and one restricted to the SARS-CoV-2 and SARS-CoV-1 clade sequences (which consists of just 12 RBD sequences). The full alignment is probably relevant to mutational constraint on expression (since we see that the Clade 2 and 3 sequencs in our homolog panel still have the same expression range as the ACE2-binding variants), while the restricted alignment is probably more relevant to mutational constraint on ACE2 binding (since the Clade 2 and 3 sequencs in our homolog panel had no detectable binding to huACE2.)

Within each of these alignments, we compute the entropy of the alignment column (and the related effective number amino acids N<sub>eff</sub>), and compare this value to the average effect of mutations at the site from our DMS data for binding and expression. We can see the sites where we observe considerable mutational constraint exhibit reduced amino acid diversity across the sarbecovirus clade and the ACE2-restricted clades in particular.

```{r natural_diversity_v_mean_effects, fig.width=7, fig.height=7, fig.align="center",dpi=300,dev="png",echo=T}
alignment <- read.fasta(file='data/alignments/RBDs_aligned.fasta')
#remove columns that is gap in the SARS-CoV-2 sequence (only one is a single A insertion in the BM48-31 sequence)
alignment$ali <- alignment$ali[,which(alignment$ali[1,]!="-")]

#get indices for alignment rows which correspond to sequences from each of the sarbecovirus 'clades'
clade1 <- which(alignment$id %in% c("SARS_Urbani_AY278741","WIV1_KF367457","WIV16_KT444582","LYRa11_KF569996",
                                    "Rs7327_KY417151","Rs4231_KY417146","RsSHC014_KC881005","Rs4084_KY417144"))

cladeSARS2 <- which(alignment$id %in% c("SARS-CoV-2_MN908947","Pangolin_GD-consensus_EPI_ISL_410544-Lam2020_supplement",
                                        "RaTG13_MN996532","Pangolin_GX-P2V_EPI_ISL_410542"))

clade2 <- which(alignment$id %in% c("As6526_KY417142","Rs4237_KY417147","Rs4081_KY417143","Rp3_DQ071615","Shaanxi2011_JX993987","279-2005_DQ648857",
                                    "Yunnan2011_JX993988","YN2013_KJ473816","Rf4092_KY417145","ZXC21_MG772934","ZC45_MG772933","JL2012_KJ473811",
                                    "Rf1_DQ412042","HeB2013_KJ473812","273-2005_DQ648856","RmYN02_EPI_ISL_412977","Rs4247_KY417148","HKU3-13_GQ153548",
                                    "HKU3-1_DQ022305","GX2013_KJ473815","Longquan-140_KF294457","HKU3-8_GQ153543","HuB2013_KJ473814"))

clade3 <- which(alignment$id %in% c("BM48-31_NC014470","BtKY72_KY352407"))

#make a second alignment which is only the clade 1 and sars-cov-2 clade sequences, where ACE2-binding is present
alignment_restricted <- alignment;alignment_restricted$ali <- alignment$ali[c(clade1,cladeSARS2),]; alignment_restricted$id <- alignment$id[c(clade1,cladeSARS2)]

entropy <- entropy(alignment)$H
entropy_restricted <- entropy(alignment_restricted)$H
RBD_sites$entropy <- entropy
RBD_sites$Neff <- exp(entropy)
RBD_sites$entropy_restricted <- entropy_restricted
RBD_sites$Neff_restricted <- exp(entropy_restricted)

#plot mean effect on binding and expression versus Neff and restricted alignment set Neff
par(mfrow=c(2,2))
plot(RBD_sites$Neff,RBD_sites$mean_bind,pch=16,col="#00000067",ylab="site mean effect of mutation on binding",xlab="site N_eff, full sarbecovirus alignment")
plot(RBD_sites$Neff,RBD_sites$mean_expr,pch=16,col="#00000067",ylab="site mean effect of mutation on expression",xlab="site N_eff, full sarbecovirus alignment")
plot(RBD_sites$Neff_restricted,RBD_sites$mean_bind,pch=16,col="#00000067",ylab="site mean effect of mutation on binding",xlab="site N_eff, SARS-CoV-1 and -2 clades")
plot(RBD_sites$Neff_restricted,RBD_sites$mean_expr,pch=16,col="#00000067",ylab="site mean effect of mutation on expression",xlab="site N_eff, SARS-CoV-1 and -2 clades")


```

## Shifting mutational preferences across sarbecovirus clades

How do the effects of individual amino acid differences among sarbecovirus homologs compare to the overall DFE? And how does this differ among clades? Below, we show the distribution of effects of amino acid differences found among sarbecovirus RBDs in each clade when that state is introduced into SARS-CoV-2. We show these distributions for all RBD sites (top), and for sites in the receptor binding motif (RBM) in particular (bottom).

Overall, we see that the amino acid variants that accumulated in each sarbecovirus clade are less deleterious for binding and expression than the overall DFE, suggesting that overall, the amino acid variation sampled in nature is subjected to some degree of purifying selection for phenotypes that correlate with our measured phenotypes -- this is particularly true for the expression phenotypes, and I imagine that the skewing in Clade2 is primarily a function of shifting amino acid preferences due to the large degree of divergence of these branches (see the relatively longer branch to this clade on our RBD phylogeny). I am actually a bit surprised to see that the amino acid states sampled in Clades 2 and 3 are not skewed toward more deleterious binding effects in the SARS-CoV-2 background. Since these have not been shown to bind ACE2s, it is unclear whether they are operating under the same selective constraint as sequences in Clade 1 and the SARS-CoV-2 clade. It is important to note that this analysis is *not* taking into account two large deletions in the RBM loops that are present in these Clade 2+3 sequences, which may also contribute to their reduced affinity for human ACE2.

```{r homolog_amino_acid_variants, fig.width=8, fig.height=8, fig.align="center",dpi=300,dev="png",echo=F}
#build up a list of amino acid variants observed in sequences from each of the clades
clade1_subs <- c()
clade1_subs_variant <- c()
clade2_subs <- c()
clade2_subs_variant <- c()
clade3_subs <- c()
clade3_subs_variant <- c()
cladeSARS2_subs <- c()
cladeSARS2_subs_variant <- c()

for(i in clade1){
  for(j in 1:ncol(alignment$ali)){
    if(alignment$ali[1,j] != alignment$ali[i,j]){
      clade1_subs <- c(clade1_subs,paste(alignment$ali[1,j],j,alignment$ali[i,j],sep=""))
      clade1_subs_variant <- c(clade1_subs_variant, alignment$id[i])
    }
  }
}
mutants_clade1 <- data.table(data.frame(mutation_RBD=clade1_subs,homolog=clade1_subs_variant))
for(i in 1:nrow(mutants_clade1)){
  if(mutants_clade1[i,mutation_RBD] %in% mutants$mutation_RBD){
    mutants_clade1[i,c("mutation","SARS2_site","bind_lib1","bind_lib2","bind_avg","expr_lib1","expr_lib2","expr_avg") := mutants[mutation_RBD==mutants_clade1[i,mutation_RBD],.(mutation,as.numeric(SARS2_site),bind_lib1,bind_lib2,bind_avg,expr_lib1,expr_lib2,expr_avg)]]
  }else{
    mutants_clade1[i,c("mutation","SARS2_site","bind_lib1","bind_lib2","bind_avg","expr_lib1","expr_lib2","expr_avg") := list(as.character(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA))]
  }
}
mutants_clade1$clade <- as.factor("clade1")

for(i in clade2){
  for(j in 1:ncol(alignment$ali)){
    if(alignment$ali[1,j] != alignment$ali[i,j]){
      clade2_subs <- c(clade2_subs,paste(alignment$ali[1,j],j,alignment$ali[i,j],sep=""))
      clade2_subs_variant <- c(clade2_subs_variant, alignment$id[i])
    }
  }
}
mutants_clade2 <- data.table(data.frame(mutation_RBD=clade2_subs,homolog=clade2_subs_variant))
for(i in 1:nrow(mutants_clade2)){
  if(mutants_clade2[i,mutation_RBD] %in% mutants$mutation_RBD){
    mutants_clade2[i,c("mutation","SARS2_site","bind_lib1","bind_lib2","bind_avg","expr_lib1","expr_lib2","expr_avg") := mutants[mutation_RBD==mutants_clade2[i,mutation_RBD],.(mutation,as.numeric(SARS2_site),bind_lib1,bind_lib2,bind_avg,expr_lib1,expr_lib2,expr_avg)]]
  }else{
    mutants_clade2[i,c("mutation","SARS2_site","bind_lib1","bind_lib2","bind_avg","expr_lib1","expr_lib2","expr_avg") := list(as.character(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA))]
  }
}
mutants_clade2$clade <- as.factor("clade2")

for(i in clade3){
  for(j in 1:ncol(alignment$ali)){
    if(alignment$ali[1,j] != alignment$ali[i,j]){
      clade3_subs <- c(clade3_subs,paste(alignment$ali[1,j],j,alignment$ali[i,j],sep=""))
      clade3_subs_variant <- c(clade3_subs_variant, alignment$id[i])
    }
  }
}
mutants_clade3 <- data.table(data.frame(mutation_RBD=clade3_subs,homolog=clade3_subs_variant))
for(i in 1:nrow(mutants_clade3)){
  if(mutants_clade3[i,mutation_RBD] %in% mutants$mutation_RBD){
    mutants_clade3[i,c("mutation","SARS2_site","bind_lib1","bind_lib2","bind_avg","expr_lib1","expr_lib2","expr_avg") := mutants[mutation_RBD==mutants_clade3[i,mutation_RBD],.(mutation,as.numeric(SARS2_site),bind_lib1,bind_lib2,bind_avg,expr_lib1,expr_lib2,expr_avg)]]
  }else{
    mutants_clade3[i,c("mutation","SARS2_site","bind_lib1","bind_lib2","bind_avg","expr_lib1","expr_lib2","expr_avg") := list(as.character(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA))]
  }
}
mutants_clade3$clade <- as.factor("clade3")

for(i in cladeSARS2){
  for(j in 1:ncol(alignment$ali)){
    if(alignment$ali[1,j] != alignment$ali[i,j]){
      cladeSARS2_subs <- c(cladeSARS2_subs,paste(alignment$ali[1,j],j,alignment$ali[i,j],sep=""))
      cladeSARS2_subs_variant <- c(cladeSARS2_subs_variant, alignment$id[i])
    }
  }
}
mutants_cladeSARS2 <- data.table(data.frame(mutation_RBD=cladeSARS2_subs,homolog=cladeSARS2_subs_variant))
for(i in 1:nrow(mutants_cladeSARS2)){
  if(mutants_cladeSARS2[i,mutation_RBD] %in% mutants$mutation_RBD){
    mutants_cladeSARS2[i,c("mutation","SARS2_site","bind_lib1","bind_lib2","bind_avg","expr_lib1","expr_lib2","expr_avg") := mutants[mutation_RBD==mutants_cladeSARS2[i,mutation_RBD],.(mutation,as.numeric(SARS2_site),bind_lib1,bind_lib2,bind_avg,expr_lib1,expr_lib2,expr_avg)]]
  }else{
    mutants_cladeSARS2[i,c("mutation","SARS2_site","bind_lib1","bind_lib2","bind_avg","expr_lib1","expr_lib2","expr_avg") := list(as.character(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA))]
  }
}
mutants_cladeSARS2$clade <- as.factor("cladeSARS2")

#put together long format table on observed substitutions
mutants_temp <- mutants[wildtype != mutation & mutation != "*",.(mutation_RBD,mutation,SARS2_site,bind_lib1,bind_lib2,bind_avg,expr_lib1,expr_lib2,expr_avg)]
mutants_temp$clade <- as.factor("full DFE"); mutants_temp$homolog <- as.factor(NA)

mutants_temp <- rbind(mutants_temp,mutants_cladeSARS2,mutants_clade1,mutants_clade2,mutants_clade3)

p1 <- ggplot(mutants_temp[!is.na(bind_avg),],aes(x=clade,y=bind_avg))+
  geom_violin(scale="width")+stat_summary(fun.y=median,geom="point",size=1)+
  xlab("clade")+ylab("delta_log10Ka,app (binding)")+theme(axis.text.x=element_text(angle=-45,hjust=0))

p2 <- ggplot(mutants_temp[!is.na(expr_avg),],aes(x=clade,y=expr_avg))+
  geom_violin(scale="width")+stat_summary(fun.y=median,geom="point",size=1)+
  xlab("clade")+ylab("delta_meanF (expression)")+theme(axis.text.x=element_text(angle=-45,hjust=0))

for(i in 1:nrow(mutants_temp)){
  if(mutants_temp$SARS2_site[i] %in% RBD_sites[RBM==T,site_SARS2]){
    mutants_temp$RBM[i] <- as.logical(TRUE)
  }else{
    mutants_temp$RBM[i] <- as.logical(FALSE)
  }
}

p3 <- ggplot(mutants_temp[!is.na(bind_avg) & RBM==T,],aes(x=clade,y=bind_avg))+
  geom_violin(scale="width")+stat_summary(fun.y=median,geom="point",size=1)+
  xlab("clade")+ylab("delta_log10Ka,app (binding)")+theme(axis.text.x=element_text(angle=-45,hjust=0))

p4 <- ggplot(mutants_temp[!is.na(expr_avg) & RBM==T,],aes(x=clade,y=expr_avg))+
  geom_violin(scale="width")+stat_summary(fun.y=median,geom="point",size=1)+
  xlab("clade")+ylab("delta_meanF (expression)")+theme(axis.text.x=element_text(angle=-45,hjust=0))

grid.arrange(p1,p2,p3,p4,nrow=2)

```

Given that ACE2-binding is restricted to Clade 1 and the SARS-CoV-2 clade, let's first focus on amino acid differences found among Clade 1 sequences (and SARS-CoV-1 in particular) and their effects in the SARS-CoV-2 background. Below are heatmaps illustrating mutational effects on binding and expression at all positions that differ in amino acid identity between SARS-CoV-1 and -2, with the "x" indicator marking the SARS-CoV-2 wildtype state, and the "o" indicator marking SARS-CoV-1 wildtype. The positions of all of the variable amino acids between these virus RBDs is shown in `dms-view` [here](https://dms-view.github.io/?pdb-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2F6m0j.pdb&markdown-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2FBloomLab_rbd.md&data-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2Fresults%2FBloomLab2020_rbd.csv&condition=natural+frequencies&site_metric=site_entropy&mutation_metric=mut_frequency&selected_sites=346%2C348%2C354%2C357%2C372%2C373%2C384%2C393%2C402%2C403%2C406%2C417%2C430%2C434%2C438%2C439%2C441%2C443%2C444%2C445%2C446%2C452%2C455%2C456%2C458%2C459%2C460%2C462%2C470%2C471%2C472%2C473%2C474%2C475%2C476%2C477%2C478%2C481%2C482%2C484%2C485%2C486%2C490%2C493%2C494%2C498%2C499%2C501%2C503%2C519%2C529), where you can see that *many* amino acid differences accumulated in the RBM and at ACE2 contact positions between these lineages.

Overall, amino acids found in SARS-CoV-1 do not have *huge* negative effects if introduced into the SARS-CoV-2 background -- but moderate affinity defects when introducing SARS-CoV-1 wildtype states with mutations K417V, L455V, II472P, A475P, G476D, G482P, F486L, and S494D, along with slight expression defects caused by S438T and N439R. These negative-effect SARS-CoV-1 mutations are highlighted in `dms-view` [here](https://dms-view.github.io/?pdb-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2F6m0j.pdb&markdown-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2FBloomLab_rbd.md&data-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2Fresults%2FBloomLab2020_rbd.csv&condition=natural+frequencies&site_metric=site_entropy&mutation_metric=mut_frequency&selected_sites=417%2C455%2C472%2C475%2C476%2C482%2C486%2C494), where we can see that they cluster in two parts of the RBM -- the lateral loop containing the RBM disulfide, and the medial portion of the RBM loop.

On the other hand, many mutations to SARS-CoV-1 amino acid identities seem to have *positive* effects on binding and expression. Most notably for expression, are mutations N354E, R357K (mild), K417V (discussed above, also in context of binding/expression trade-off), and L452K, and for binding, beneficial effects are also seen for N460K, Q498V (quite substantial positive effect, direct interface residue), N501T (also quite substantial, interface, site of key adaptation in SARS-CoV-1), V503I, and H519N. These positions are highlighted on the structure [here](https://dms-view.github.io/?pdb-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2F6m0j.pdb&markdown-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2FBloomLab_rbd.md&data-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2Fresults%2FBloomLab2020_rbd.csv&condition=natural+frequencies&site_metric=site_entropy&mutation_metric=mut_frequency&selected_sites=354%2C357%2C417%2C452%2C460%2C498%2C501%2C503%2C519), where we can see that many are distant from the ACE2 interface (primarily the expression-enhancing mutations). Interestingly, the beneficial *binding* SARS-CoV-1 residues are on the *other* lateral end of the RBM compared to the detrimental swaps highlighted above. Therefore, it seems like, of the interface differences between SARS-CoV-1 and -2, one "edge" of the RBM saddle could be better optimized for ACE2-binding by mutating to the SARS-CoV-1 states, while the middle and the other edge are not able to be swapped. As we will begin to see later on, the SARS-CoV-1 affinity is *not* predicted by the sum of effects of these contact mutations, which would predict a substantially worse affinity. Therefore, this observation about the deleterious SARS-CoV-1 swaps in the SARS-CoV-2 background is a reflection of epistasis among the many RBM differences that separate these two molecules. I am actually surprised to see that there are seemingly so many *beneficial* swaps to SARS-CoV-1 identities, in light of the fact that SARS-CoV-2 has repeatedly been demonstrated to have overall tighter ACE2-binding affinity than SARS-CoV-1.

```{r heatmap_SARS_CoV_2_1_diff, fig.width=12,fig.height=8,fig.align="center", dpi=500,dev="png",echo=FALSE}
#order mutant as a factor for grouping by rough biochemical grouping
mutants$mutant <- factor(mutants$mutant, levels=c("*","C","P","G","V","M","L","I","A","F","W","Y","T","S","N","Q","E","D","H","K","R"))
#add character vector indicating wildtype SARS-CoV-2 aa to use as plotting symbols for wt
mutants[,wildtype_indicator := ""]
mutants[mutant==wildtype,wildtype_indicator := "x"]

#indicator for wildtype SARS-CoV-1 state
mutants[,SARS1_indicator := ""]
for(i in 1:nrow(mutants)){
  SARS1aa <- RBD_sites[site_SARS2==mutants$SARS2_site[i],amino_acid_SARS1]
  if(!is.na(SARS1aa) & mutants$mutant[i] == SARS1aa){mutants$SARS1_indicator[i] <- "o"}
}

muts_temp <- mutants[SARS2_site %in% RBD_sites[amino_acid_SARS1 != amino_acid_SARS2,site_SARS2],];muts_temp$SARS2_site <- as.factor(muts_temp$SARS2_site)

p1 <- ggplot(muts_temp,aes(SARS2_site,mutant))+geom_tile(aes(fill=bind_avg))+
  scale_fill_gradientn(colours=c("#A94E35","#F48365","#FFFFFF","#7378B9","#383C6C"),limits=c(-5,0.5),values=c(0,2.5/5.5,5/5.5,5.25/5.5,5.5/5.5),na.value="gray")+
  scale_x_discrete(expand=c(0,0))+
  labs(x="RBD site",y="mutant")+theme_classic(base_size=9)+
  coord_equal()+theme(axis.text.x = element_text(angle=45,hjust=1))+
  geom_text(aes(label=wildtype_indicator),size=2,color="gray10")+
  geom_text(aes(label=SARS1_indicator),size=2,color="gray10")

p2 <- ggplot(muts_temp,aes(SARS2_site,mutant))+geom_tile(aes(fill=expr_avg))+
  scale_fill_gradientn(colours=c("#A94E35","#A94E35","#F48365","#FFFFFF","#7378B9","#383C6C"),limits=c(-5,1),values=c(0,1/6,3/6,5/6,5.5/6,6/6),na.value="gray")+
  scale_x_discrete(expand=c(0,0))+
  labs(x="RBD site",y="mutant")+theme_classic(base_size=9)+
  coord_equal()+theme(axis.text.x = element_text(angle=45,hjust=1))+
  geom_text(aes(label=wildtype_indicator),size=2,color="gray10")+
  geom_text(aes(label=SARS1_indicator),size=2,color="gray10")

grid.arrange(p1,p2,ncol=1)

invisible(dev.print(pdf, paste(config$sarbecovirus_diversity_dir,"/heatmaps_SARS-2-1-differences.pdf",sep="")))

```

Let's output a table of all Clade 1 amino acid variants that have moderately large detrimental effects in the SARS-CoV-2 background. This focuses the observations made from the heatmap below, and shows their broader context in Clade 1 isolate sequences. Of note, L455 and S494 are "key adaptations" in SARS-CoV-1 (Y442 and D480), according to studies by e.g. Fang Li -- so, should be a lot of prior structural work we can look at to better understand what other RBM/interface residues might be impacting amino acid preferences at these sites between SARS-CoV-1 and 2. (And perhaps suggests a future mutational scan in the SARS-CoV RBM would illustrate lots of shifting preferences!)

```{r table_del_states_clade_1, echo=F}
kable(mutants_temp[clade %in% c("clade1") & bind_avg < -1,.(mutation,bind_lib1,bind_lib2,bind_avg,expr_avg,clade,homolog)])
```

## Sequence and functional variation within the SARS-CoV-2 clade

Although few sequences have been isolated from the SARS-CoV-2 clade, we uncovered functional diversity, with GD-Pangolin exhibiting slightly enhanced ACE2-binding affinity, and RaTG13 exhibiting strongly reduced affinity. Let's take a look at the amino acid variation among these strains and the effects of these mutations in SARS-CoV-2.

First, we illustrate heatmaps for positions that differ between RaTG13 and SARS-CoV-2. The "x" indicator continues to mark the SARS-CoV-2 state, with "^" indicating the RaTG13 state. The main thing to note, is that site 501 seems to explain a large degree of difference between RaTG13 and SARS-CoV-2 affinity. We can do preliminary "by eye" parsimony reconstructions of the ancestral RaTG13+SARS-CoV2 state, which suggests that the N501D substitution is a derived state in the RaTG13 lineage (so, not as I first thought, a pointer toward adaptation on the lineage toward SARS-CoV-2). Site 501 was a key site of adaptation between civet- and human-adapted strains in SARS-CoV-1, so it's interesting to see it cropping up again here, regardless of the polarization of the change (and different amino acid states than the S/T SARS-CoV-1 stuff). Mutations to the RaTG13 amino acid are also mildly deleterious w.r.t. binding for R403T (GD-Pangolin has R, GX-Pangolin K), Y449F (GD-Pangolin has Y, GX-Pangolin Y), F486L (GD-Pangolin has F, GX-Pangolin L...), and Y505H (GD-Pangolin has Y, GX-Pangolin Y). So, from parsimony, it does seem most of these 'deleterious' amino acids are *derived* substitutions on the RaTG13 lineage, instead of reflecting 'adaptive' changes on the SARS-CoV-2 ancestral lineage.

```{r heatmap_SARS_CoV_2_RaTG13_diff, fig.width=10,fig.height=5,fig.align="center", dpi=500,dev="png",echo=FALSE}
#indicator for wildtype RaTG13 state
mutants[,RaTG13_indicator := ""]
for(i in 1:nrow(mutants)){
  RaTG13aa <- RBD_sites[site_SARS2==mutants$SARS2_site[i],amino_acid_RaTG13]
  if(!is.na(RaTG13aa) & mutants$mutant[i] == RaTG13aa){mutants$RaTG13_indicator[i] <- "^"}
}

muts_temp <- mutants[SARS2_site %in% RBD_sites[amino_acid_RaTG13 != amino_acid_SARS2,site_SARS2],];muts_temp$SARS2_site <- as.factor(muts_temp$SARS2_site)

p1 <- ggplot(muts_temp,aes(SARS2_site,mutant))+geom_tile(aes(fill=bind_avg))+
  scale_fill_gradientn(colours=c("#A94E35","#F48365","#FFFFFF","#7378B9","#383C6C"),limits=c(-5,0.5),values=c(0,2.5/5.5,5/5.5,5.25/5.5,5.5/5.5),na.value="gray")+
  scale_x_discrete(expand=c(0,0))+
  labs(x="RBD site",y="mutant")+theme_classic(base_size=9)+
  coord_equal()+theme(axis.text.x = element_text(angle=45,hjust=1))+
  geom_text(aes(label=wildtype_indicator),size=2,color="gray10")+
  geom_text(aes(label=RaTG13_indicator),size=3,color="gray10")

p2 <- ggplot(muts_temp,aes(SARS2_site,mutant))+geom_tile(aes(fill=expr_avg))+
  scale_fill_gradientn(colours=c("#A94E35","#A94E35","#F48365","#FFFFFF","#7378B9","#383C6C"),limits=c(-5,1),values=c(0,1/6,3/6,5/6,5.5/6,6/6),na.value="gray")+
  scale_x_discrete(expand=c(0,0))+
  labs(x="RBD site",y="mutant")+theme_classic(base_size=9)+
  coord_equal()+theme(axis.text.x = element_text(angle=45,hjust=1))+
  geom_text(aes(label=wildtype_indicator),size=2,color="gray10")+
  geom_text(aes(label=RaTG13_indicator),size=3,color="gray10")

grid.arrange(p1,p2,ncol=2)

invisible(dev.print(pdf, paste(config$sarbecovirus_diversity_dir,"/heatmaps_SARS-2-RaTG13-differences.pdf",sep="")))

```

Last, let's look at heatmaps for positions that differ between the GD-Pangolin and SARS-CoV-2 RBDs. We can see that 3 of the 6 differences are predicted to enhance SARS-CoV-2 affinity (2 of these also enhance expression, and 2 binding-neutral mutations also enhance expressoin), likely explaining the observed enhanced GD-Pangolin affinity. These six sites are visualized on the structure in `dms-view` [here](https://dms-view.github.io/?pdb-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2F6m0j.pdb&markdown-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2FBloomLab_rbd.md&data-url=https%3A%2F%2Fraw.githubusercontent.com%2Fdms-view%2FSARS-CoV-2%2Fmaster%2Fdata%2FSpike%2FBloomLab2020%2Fresults%2FBloomLab2020_rbd.csv&condition=natural+frequencies&site_metric=site_entropy&mutation_metric=mut_frequency&selected_sites=346%2C372%2C402%2C417%2C498%2C519).

```{r heatmap_SARS_CoV_2_GDPangolin_diff, fig.width=5,fig.height=4,fig.align="center", dpi=500,dev="png",echo=FALSE}
#indicator for wildtype GD-Pangolin state
mutants[,GDPangolin_indicator := ""]
for(i in 1:nrow(mutants)){
  GDPangolinaa <- RBD_sites[site_SARS2==mutants$SARS2_site[i],amino_acid_GD_Pangolin]
  if(!is.na(GDPangolinaa) & mutants$mutant[i] == GDPangolinaa){mutants$GDPangolin_indicator[i] <- "#"}
}

muts_temp <- mutants[SARS2_site %in% RBD_sites[amino_acid_GD_Pangolin != amino_acid_SARS2,site_SARS2],];muts_temp$SARS2_site <- as.factor(muts_temp$SARS2_site)

p1 <- ggplot(muts_temp,aes(SARS2_site,mutant))+geom_tile(aes(fill=bind_avg))+
  scale_fill_gradientn(colours=c("#A94E35","#F48365","#FFFFFF","#7378B9","#383C6C"),limits=c(-5,0.5),values=c(0,2.5/5.5,5/5.5,5.25/5.5,5.5/5.5),na.value="gray")+
  scale_x_discrete(expand=c(0,0))+
  labs(x="RBD site",y="mutant")+theme_classic(base_size=9)+
  coord_equal()+theme(axis.text.x = element_text(angle=45,hjust=1))+
  geom_text(aes(label=wildtype_indicator),size=2,color="gray10")+
  geom_text(aes(label=GDPangolin_indicator),size=2,color="gray10")

p2 <- ggplot(muts_temp,aes(SARS2_site,mutant))+geom_tile(aes(fill=expr_avg))+
  scale_fill_gradientn(colours=c("#A94E35","#A94E35","#F48365","#FFFFFF","#7378B9","#383C6C"),limits=c(-5,1),values=c(0,1/6,3/6,5/6,5.5/6,6/6),na.value="gray")+
  scale_x_discrete(expand=c(0,0))+
  labs(x="RBD site",y="mutant")+theme_classic(base_size=9)+
  coord_equal()+theme(axis.text.x = element_text(angle=45,hjust=1))+
  geom_text(aes(label=wildtype_indicator),size=2,color="gray10")+
  geom_text(aes(label=GDPangolin_indicator),size=2,color="gray10")

grid.arrange(p1,p2,ncol=2)

invisible(dev.print(pdf, paste(config$sarbecovirus_diversity_dir,"/heatmaps_SARS-2-GDPangolin-differences.pdf",sep="")))

```

## Predicting (or failing to predict) homolog affinities from single mutant effect measurements

How well does the addition of single mutation effects in the SARS-CoV-2 background predict homolog affinities? We try this in three ways: 1) summing mutational effects across all RBD positions, 2) just RBM positions, 3) just ACE2 contact positions. Expectation is: predictions of RaTG13 and pangolin affinity might be decent, but clade1 and more distant sequences will be bad. (We can also propagate the n=2 SEM for these additions, in case useful.) The table below gives the number of differences between the sequence and SARS-CoV-2 in each of the query classes. It then compares the experimental binding and expression scores (bind_ and expr_avg) to those predicted by summing the mutational effects of individual differences in these three categories. 

We can see that the relative predictions for RaTG13 and GD-Pangolin are correct, though the magnitude of affinity decrease in RaTG13 is over-predicted. For Clade 1 sequences, we can see that the overall magnitude of affinity decrease is severely over-predicted -- but among the Clade 1 sequences, the relative affinity order isn't terrible. That is, we do predict that WIV16 has higher affinity than SARS-CoV-1 or LYRa11, though we don't also successfully predict that SARS-CoV-1 has higher affinity than LYRa11. Clade 2 sequences are predicted to have universally terrible binding, as we'd expect. And surprisingly, BM48-31 is not predicted to have terrible binding. I interpret this to mean that, although neither Clade 2 or BM48-31 bind ACE2, BM48-31 is genetically "closer" to ACE2-binding, insofar as, it has fewer amino acid differences that are measured as being highly incompatible with ACE2-binding. (Alternative is that BM48-31 is less distinct from SARS-CoV-2 at the level of shifted amino acid preference, so its fixed states aren't interpreted as being as incompatible as those in Clade 2 -- this is consistent with the observation that BM48-31 is predicted to be better expressing than Clade 2 sequences, as well, reflecting overall less divergent amino acid preferences.) Either of these scenarios are perhaps supported by an observation from Letko et al. 2020 in [Figure 4](https://www.nature.com/articles/s41564-020-0688-y/figures/4) -- in this experiment, they made chimeras of Clade 2 and Clade 3 core RBDs with the Clade 1 RBM, and they saw that the Clade 3 core RBD supported higher cellular infectivity than Clade 2 when grafted with the appropriate RBM.

```{r predict_homolog_affinities, echo=F}
mutants[,bind_SEM := sd(c(bind_lib1,bind_lib2))/sqrt(2),by=mutation]
mutants[,expr_SEM := sd(c(expr_lib1,expr_lib2))/sqrt(2),by=mutation]

homolog_measurements$alignment_name <- c("SARS-CoV-2_MN908947","Pangolin_GD-consensus_EPI_ISL_410544-Lam2020_supplement","RaTG13_MN996532","SARS_Urbani_AY278741","WIV16_KT444582",
                                         "LYRa11_KF569996","ZC45_MG772933","ZXC21_MG772934","HKU3-1_DQ022305","Rf1_DQ412042","Rp3_DQ071615","BM48-31_NC014470")

for(i in 1:nrow(homolog_measurements)){
  subs <- c()
  for(j in 1:ncol(alignment$ali)){
    if(!(alignment$ali[which(alignment$id==homolog_measurements[i,alignment_name]),j] %in% c("-","X")) & alignment$ali[1,j] != alignment$ali[which(alignment$id==homolog_measurements[i,alignment_name]),j]){
      subs <- c(subs,paste(alignment$ali[1,j],j,alignment$ali[which(alignment$id==homolog_measurements[i,alignment_name]),j],sep=""))
    }
  }
  homolog_measurements$n_diff[i] <- length(subs)
  homolog_measurements$sum_bind[i] <- sum(mutants[mutation_RBD %in% subs,bind_avg])
  homolog_measurements$SEM_sum_bind[i] <- sqrt(sum(mutants[mutation_RBD %in% subs,bind_SEM]^2))
  homolog_measurements$sum_expr[i] <- sum(mutants[mutation_RBD %in% subs,expr_avg])
  homolog_measurements$SEM_sum_expr[i] <- sqrt(sum(mutants[mutation_RBD %in% subs,expr_SEM]^2))
}

for(i in 1:nrow(homolog_measurements)){
  subs <- c()
  for(j in RBD_sites[RBM==T,site_RBD]){
    if(!(alignment$ali[which(alignment$id==homolog_measurements[i,alignment_name]),j] %in% c("-","X")) & alignment$ali[1,j] != alignment$ali[which(alignment$id==homolog_measurements[i,alignment_name]),j]){
      subs <- c(subs,paste(alignment$ali[1,j],j,alignment$ali[which(alignment$id==homolog_measurements[i,alignment_name]),j],sep=""))
    }
  }
  homolog_measurements$n_diff_RBM[i] <- length(subs)
  homolog_measurements$sum_bind_RBM[i] <- sum(mutants[mutation_RBD %in% subs,bind_avg])
  homolog_measurements$SEM_sum_bind_RBM[i] <- sqrt(sum(mutants[mutation_RBD %in% subs,bind_SEM]^2))
  homolog_measurements$sum_expr_RBM[i] <- sum(mutants[mutation_RBD %in% subs,expr_avg])
  homolog_measurements$SEM_sum_expr_RBM[i] <- sqrt(sum(mutants[mutation_RBD %in% subs,expr_SEM]^2))
}

for(i in 1:nrow(homolog_measurements)){
  subs <- c()
  for(j in RBD_sites[SARS2_ACE2_contact==T | SARS1_ACE2_contact==T,site_RBD]){
    if(!(alignment$ali[which(alignment$id==homolog_measurements[i,alignment_name]),j] %in% c("-","X")) & alignment$ali[1,j] != alignment$ali[which(alignment$id==homolog_measurements[i,alignment_name]),j]){
      subs <- c(subs,paste(alignment$ali[1,j],j,alignment$ali[which(alignment$id==homolog_measurements[i,alignment_name]),j],sep=""))
    }
  }
  homolog_measurements$n_diff_contact[i] <- length(subs)
  homolog_measurements$sum_bind_contact[i] <- sum(mutants[mutation_RBD %in% subs,bind_avg])
  homolog_measurements$SEM_sum_bind_contact[i] <- sqrt(sum(mutants[mutation_RBD %in% subs,bind_SEM]^2))
  homolog_measurements$sum_expr_contact[i] <- sum(mutants[mutation_RBD %in% subs,expr_avg])
  homolog_measurements$SEM_sum_expr_contact[i] <- sqrt(sum(mutants[mutation_RBD %in% subs,expr_SEM]^2))
}

kable(homolog_measurements[,.(homolog,n_diff,n_diff_RBM,n_diff_contact,bind_avg,sum_bind,sum_bind_RBM,sum_bind_contact,expr_avg,sum_expr,sum_expr_RBM,sum_expr_contact)],
      col.names=c("homolog","# RBD diff","# RBM diff","# contact diff","measured binding","predicted binding, RBD","predicted binding, RBM","predicted binding, contact","measured expression", "predicted expression, all RBD","predicted expression, RBM","predicted expression, contact"))

```

To visualize the trends in this table, we plot the sum of binding and expression effects versus the number of mutations present in a genotype for RBM residues. Once again, this does not take into account deletions within RBM loops present in Clade2 and Clade3 sequences. We then plot the sum of mutational effects on binding versus expression, which could potentially "normalize" out the extent of shifting preferences, because all variants express similarly well. Finally, we plot the actual measured binding versus predicted by summing mutational effects. Points are colored by clade: purple (blue?) is SARS-CoV-2 clade sequences, orange (red?) is Clade 1 (SARS-CoV-1 clade), green is Clade 2 (deletions, have not been seen to use ACE2), magenta is clade 3 (Bulgaria and Kenya isolates) (sorry my descriptions of colors are probably wrong :/ let me know if the color scheme is hideous, I've been known to pick terrible colors in the past -- these are the colors I included in our phylogeny figure so we should update all of these ASAP if needed)

The first trend this shows is that the number of differences is a primary determinant of how "deleterious" a sequence is predicted to be -- both for binding and expression -- which probably reflects epistasis and shifting preferences in the more distant backgrounds as I was starting to say above the table. So, it might be that these predictions work ok for within the SARS-CoV-2 clade where the number of differences is relatively small -- but in other clades, the number of differences is just too large for the sum to be relevant. However, maybe within these other clades, the *relative* binding predicted for one e.g. clade 1 sequence versus another predicted is useful.

```{r predicted_versus_actual_homolog_correlations_RBM, fig.width=7, fig.height=7, fig.align="center",dpi=300,dev="png",echo=F}
par(mfrow=c(2,2))
plot(homolog_measurements$n_diff_RBM,homolog_measurements$sum_bind_RBM,col=homolog_measurements$clade_color,pch=16,xlab="number differences (RBM residues)",ylab="sum of mutational effects on binding",cex=2)
plot(homolog_measurements$n_diff_RBM,homolog_measurements$sum_expr_RBM,col=homolog_measurements$clade_color,pch=16,xlab="number differences (RBM residues)",ylab="sum of mutational effects on expression",cex=2)

plot(homolog_measurements$sum_expr_RBM,homolog_measurements$sum_bind_RBM,col=homolog_measurements$clade_color,pch=16,xlab="sum of mutational effects on expression",ylab="sum of mutational effects on binding",cex=2)
plot(homolog_measurements$bind_avg,homolog_measurements$sum_bind_RBM,col=homolog_measurements$clade_color,pch=16,xlab="experimentally determined binding",ylab="sum of mutational effects on binding",cex=2)

```
And the same plots, for only ACE2-contact residue positions:

```{r predicted_versus_actual_homolog_correlations_contact, fig.width=7, fig.height=7, fig.align="center",dpi=300,dev="png",echo=F}
par(mfrow=c(2,2))
plot(homolog_measurements$n_diff_contact,homolog_measurements$sum_bind_contact,col=homolog_measurements$clade_color,pch=16,xlab="number differences (contact residues)",ylab="sum of mutational effects on binding",cex=2)
plot(homolog_measurements$n_diff_contact,homolog_measurements$sum_expr_contact,col=homolog_measurements$clade_color,pch=16,xlab="number differences (contact residues)",ylab="sum of mutational effects on expression",cex=2)

plot(homolog_measurements$sum_expr_contact,homolog_measurements$sum_bind_contact,col=homolog_measurements$clade_color,pch=16,xlab="sum of mutational effects on expression",ylab="sum of mutational effects on binding",cex=2)
plot(homolog_measurements$bind_avg,homolog_measurements$sum_bind_contact,col=homolog_measurements$clade_color,pch=16,xlab="experimentally determined binding",ylab="sum of mutational effects on binding",cex=2)

```

(What does it look like when we add in the other Clade 1/2/3 sequences that we did not measure, but see where they fall on the predicted binding versus expression plots? Particularly GX Pangolin)

Overall, my intpretation from this table/figures, is that our mutational effects are perhaps useful for predictions of things very similiar to our focal background (e.g. GD Pangolin, circulating strains), but that because of epistasis, mutational preferences are quite different in other sarbecovirus clades, which complicates prediction of ability of unseen RBDs to bind ACE2. (This perhaps can mentally be loosely accounted for by taking into account the total number of amino acid differences and also their effect on expression, but that's a lot of weirdness to normalize for so overall definitely not a direction to push.) These are probably not the figures we'll want to use to communicate this, although I do think this is an interestning conclusion to communicate (that preferences have shifted between SARS-CoV-2 and -1)


## Epistasis in library double mutants and relation to natural diversity patterns

*Note* though I do see some structurally interesting epistasis when analyzing double mutant barcodes compared to the effects of their component single mutations, I didn't really see anything interesting below as it relates to the goal I intended toward with putting this anlaysis in this script (seeing if I could see pairwise epistatic modifiers for the sites/states observed in SARS-CoV-1 that are not well tolerated in SARS-CoV-2). Therefore, I've moved the "structurally relevant" component of this script to the structure_function notebook, where it might be elaborated in that context. I keep some remaining code chunkd hidden here related to what I tried w.r.t. natural diversity in case I decide to come back to it, but this will probably eventually be eliminated (and doesn't work right now as the upstream analysis before making these plots is now removed into a different notebook)

```{r positive_epistasis_Clade1_incompatibilities_scatters, fig.width=6, fig.height=6, fig.align="center",dpi=300,dev="png",echo=F}
# par(mfrow=c(2,2))
# 
# plot(bc_dbl[(site1 == 455 | site2 == 455),mut1_bind] + bc_dbl[(site1 == 455 | site2 == 455),mut2_bind], bc_dbl[(site1 == 455 | site2 == 455),delta_log10Ka],xlab="sum of component single mutations",ylab="double mutant barcode direct phenotype",pch=16,col="#00000090",main="site 455 double mutants")
# points(bc_dbl[(mut1 == "L455Y" | mut2 == "L455Y"),mut1_bind] + bc_dbl[(mut1 == "L455Y" | mut2 == "L455Y"),mut2_bind], bc_dbl[(mut1 == "L455Y" | mut2 == "L455Y"),delta_log10Ka],col="green",pch=16,cex=1.3)
# abline(a=0,b=1,col="red",lty=2,lwd=2)
# 
# plot(bc_dbl[(site1 == 475 | site2 == 475),mut1_bind] + bc_dbl[(site1 == 475 | site2 == 475),mut2_bind], bc_dbl[(site1 == 475 | site2 == 475),delta_log10Ka],xlab="sum of component single mutations",ylab="double mutant barcode direct phenotype",pch=16,col="#00000090",main="site 475 double mutants")
# points(bc_dbl[(mut1 == "A475P" | mut2 == "A475P"),mut1_bind] + bc_dbl[(mut1 == "A475P" | mut2 == "A475P"),mut2_bind], bc_dbl[(mut1 == "A475P" | mut2 == "A475P"),delta_log10Ka],col="green",pch=16,cex=1.3)
# abline(a=0,b=1,col="red",lty=2,lwd=2)
# 
# plot(bc_dbl[(site1 == 494 | site2 == 494),mut1_bind] + bc_dbl[(site1 == 494 | site2 == 494),mut2_bind], bc_dbl[(site1 == 494 | site2 == 494),delta_log10Ka],xlab="sum of component single mutations",ylab="double mutant barcode direct phenotype",pch=16,col="#00000090",main="site 494 double mutants")
# points(bc_dbl[(mut1 == "S494D" | mut2 == "S494D"),mut1_bind] + bc_dbl[(mut1 == "S494D" | mut2 == "S494D"),mut2_bind], bc_dbl[(mut1 == "S494D" | mut2 == "S494D"),delta_log10Ka],col="green",pch=16,cex=1.3)
# abline(a=0,b=1,col="red",lty=2,lwd=2)
# 
# plot(bc_dbl[(site1 == 449 | site2 == 449),mut1_bind] + bc_dbl[(site1 == 449 | site2 == 449),mut2_bind], bc_dbl[(site1 == 449 | site2 == 449),delta_log10Ka],xlab="sum of component single mutations",ylab="double mutant barcode direct phenotype",pch=16,col="#00000090",main="site 449 double mutants")
# points(bc_dbl[(mut1 == "Y449F" | mut2 == "Y449F"),mut1_bind] + bc_dbl[(mut1 == "Y449F" | mut2 == "Y449F"),mut2_bind], bc_dbl[(mut1 == "Y449F" | mut2 == "Y449F"),delta_log10Ka],col="green",pch=16,cex=1.3)
# abline(a=0,b=1,col="red",lty=2,lwd=2)
```

```{r double_mut_bcs_A475P, echo=F}
# #table sorted by positive epistasis score for position 475
# bc_dbl[(site1 == 475 | site2 == 475) & delta_log10Ka > -4, .(avgcount,mut1,mut2,CA_dist,mut1_bind,mut2_bind,delta_log10Ka,epistasis_bind,mut1_expr,mut2_expr,delta_ML_meanF)][order(epistasis_bind,decreasing=T),]
# #table sorted by positive epistasis score for mutation A475P
# bc_dbl[(mut1 == "A475P" | mut2 == "A475P"), .(avgcount,mut1,mut2,CA_dist,mut1_bind,mut2_bind,delta_log10Ka,epistasis_bind,mut1_expr,mut2_expr,delta_ML_meanF)][order(epistasis_bind,decreasing=T),]
```

```{r double_mut_bcs_S494D, echo=F}
# #table sorted by positive epistasis score for position 475
# bc_dbl[(site1 == 494 | site2 == 494) & delta_log10Ka > -4, .(avgcount,mut1,mut2,CA_dist,mut1_bind,mut2_bind,delta_log10Ka,epistasis_bind,mut1_expr,mut2_expr,delta_ML_meanF)][order(epistasis_bind,decreasing=T),]
# #table sorted by positive epistasis score for mutation A475P
# bc_dbl[(mut1 == "S494D" | mut2 == "S494D"), .(avgcount,mut1,mut2,CA_dist,mut1_bind,mut2_bind,delta_log10Ka,epistasis_bind,mut1_expr,mut2_expr,delta_ML_meanF)][order(epistasis_bind,decreasing=T),]
```

```{r double_mut_bcs_Y449F, echo=F}
# #table sorted by positive epistasis score for position 475
# bc_dbl[(site1 == 449 | site2 == 449) & delta_log10Ka > -4, .(avgcount,mut1,mut2,CA_dist,mut1_bind,mut2_bind,delta_log10Ka,epistasis_bind,mut1_expr,mut2_expr,delta_ML_meanF)][order(epistasis_bind,decreasing=T),]
# #table sorted by positive epistasis score for mutation A475P
# bc_dbl[(mut1 == "Y449F" | mut2 == "Y449F"), .(avgcount,mut1,mut2,CA_dist,mut1_bind,mut2_bind,delta_log10Ka,epistasis_bind,mut1_expr,mut2_expr,delta_ML_meanF)][order(epistasis_bind,decreasing=T),]
```



