---
title: "Output MaveDB uploads"
author: "Tyler Starr"
date: "6/18/2020"
output:
  github_document:
    toc: true
    html_preview: false
editor_options:
  chunk_output_type: inline
---

This notebook outputs DMS measurements in the format required for upload to [MaveDB](https://www.mavedb.org)

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

#list of packages to install/load
packages = c("yaml","data.table","tidyverse","seqinr")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages])
}
#load packages
invisible(lapply(packages, library, character.only=T))

#read in config file
config <- read_yaml("../config.yaml")

```

## MaveDB organization

I defined an "Experiment Set" on MaveDB, which contains two "Experiments": "Effects on binding of mutation in the SARS-CoV-2 RBD" and "Effects on expression of mutations in the SARS-CoV-2 RBD". Within each of these experiments, we will upload two "Score Sets", consisting of our measurements (delta-log<sub>10</sub>(_K_<sub>D,app</sub>) for binding and delta-log(MFI) for expression), at the level of individual barcodes, and our final single-mutation-effect dataset.

## Per-barcode score set generation

Let's read in the per-barcode binding and expression measurements.

```{r bc_bind_input}
bc_bind <- data.table(read.csv(file=paste("../",config$Titeseq_Kds_file,sep="")))
head(bc_bind)
```

```{r bc_expr_input}
bc_expr <- data.table(read.csv(file=paste("../",config$expression_sortseq_file,sep="")))
head(bc_expr)
```
We need to rename the genotype column to `hgvs_pro` and conform to [HGVS guidelines](http://varnomen.hgvs.org/recommendations/general/). For protein mutations, we:

   - Count from the first mutated position as position 1 (I believe we can add the offset for spike numbering later on)
   - Convert to three-letter amino acid codes
   - For single mutants, add the `p.` designator to specify amino acid changes, e.g. `p.Asn1Gln`
   - For multiple mutants, format as `p.[Asn1Gln;Ile2Leu]`
   - Stop mutants stay as single letter `*` character

We then rename our functional score for each table to be `score`

We include as optional columns for the binding data:

   - `avg_count` (currently named `avgcount`), `barcode`, and `library`

We include as optional columns for the expression data:

  - `count` (currently named `total_count`), `barcode`, and `library`

Below, we make these modifications and output the resulting `.csv` files

```{r bc_bind_output}
#modify aaa() function to not modify * to Stp, as HGVS still prefers * for stop even in three letter code
aaa_stop <- function(x){if(x=="*"){return(x)}else{return(aaa(x))}}

#function to return HGVS nomenclature mutants from our one letter code notation
hgvs_naming <- function(x){
  if(x==""){
    return("")
  }else{
    muts <- strsplit(as.character(x),split=" ")[[1]]
    if(length(muts)==1){
      split <- strsplit(muts,split="")[[1]]
      split[1] <- aaa_stop(split[1])
      split[length(split)] <- aaa_stop(split[length(split)])
      paste <- paste("p.",paste(split,collapse=""),sep="")
      return(paste)
    }else if(length(muts) > 1){
      for(i in 1:length(muts)){
        split <- strsplit(muts[i],split="")[[1]]
        split[1] <- aaa_stop(split[1])
        split[length(split)] <- aaa_stop(split[length(split)])
        paste <- paste(split,collapse="")
        muts[i] <- paste
      }
      paste("p.[",paste(muts, collapse=";"),"]",sep="")
    }
  }
}

bc_bind[, hgvs_pro := hgvs_naming(aa_substitutions),by="aa_substitutions"]

setnames(bc_bind, "delta_log10Ka", "score"); setnames(bc_bind, "avgcount", "avg_count")

bc_bind <- bc_bind[,.(hgvs_pro, score, avg_count, library)]

bc_bind[library=="lib1",library:="1"]
bc_bind[library=="lib2",library:="2"]

#empty unmutated sequence not allowed. Try "p.="
bc_bind[hgvs_pro=="",hgvs_pro:="p.="]

head(bc_bind)

write.csv(bc_bind,file="score-set_binding_bc.csv",row.names=F)
```

```{r bc_expr_output}
bc_expr[, hgvs_pro := hgvs_naming(aa_substitutions),by="aa_substitutions"]

setnames(bc_expr, "delta_ML_meanF", "score"); setnames(bc_expr, "total_count", "count")

bc_expr <- bc_expr[,.(hgvs_pro, score, count, library)]

bc_expr[library=="lib1",library:="1"]
bc_expr[library=="lib2",library:="2"]

#empty unmutated sequence not allowed. Try "p.="
bc_expr[hgvs_pro=="",hgvs_pro:="p.="]

head(bc_expr)

write.csv(bc_expr,file="score-set_expression_bc.csv",row.names=F)
```

## Per-mutation score set generation

We'll perform similar actions to output score sets consisting of our final estimates of single-mutant effects.

```{r mut_bind_input}
muts <- data.table(read.csv(file=paste("../",config$single_mut_effects_file,sep=""),stringsAsFactors=F))
muts <- muts[mutant != wildtype,]
head(muts)
```

We need to rename the `mutation_RBD` column to `hgvs_pro` and conform to [HGVS guidelines](http://varnomen.hgvs.org/recommendations/general/). For protein mutations, we:

   - Count from the first mutated position as position 1 (I believe we can add the offset for spike numbering later on)
   - Convert to three-letter amino acid codes
   - Add the `p.` designator to specify amino acid changes, e.g. `p.Asn1Gln`
   - Stop mutants stay as single letter `*` character

We then split the table into binding and expression tables, and rename our functional score for each table to be `score`

We include as optional columns for the both datasets:

   - `library` (so, collapse to long form instead of wide), and we'll also include the `average` with each row as the average of the two measurements

Below, we make these modifications and output the resulting `.csv` files

```{r mut_output}
muts[,hgvs_pro := paste("p.",aaa_stop(wildtype),site_RBD,aaa_stop(mutant),sep=""),by=mutation]
mut_expr <- muts

mut_bind_lib1 <- muts[,.(hgvs_pro,bind_lib1,bind_avg)]
mut_bind_lib1[,library:="1"]
mut_bind_lib2 <- muts[,.(hgvs_pro,bind_lib2,bind_avg)]
mut_bind_lib2[,library:="2"]

mut_expr_lib1 <- muts[,.(hgvs_pro,expr_lib1,expr_avg)]
mut_expr_lib1[,library:="1"]
mut_expr_lib2 <- muts[,.(hgvs_pro,expr_lib2,expr_avg)]
mut_expr_lib2[,library:="2"]

setnames(mut_bind_lib1, c("bind_lib1", "bind_avg"), c("score", "average"))
setnames(mut_bind_lib2, c("bind_lib2", "bind_avg"), c("score", "average"))
setnames(mut_expr_lib1, c("expr_lib1", "expr_avg"), c("score", "average"))
setnames(mut_expr_lib2, c("expr_lib2", "expr_avg"), c("score", "average"))

mut_bind <- rbind(mut_bind_lib1, mut_bind_lib2)

head(mut_bind)

write.csv(mut_bind,file="score-set_binding_mut.csv",row.names=F)
```

```{r mut_expr_output}
mut_expr <- rbind(mut_expr_lib1, mut_expr_lib2)

head(mut_expr)

write.csv(mut_expr,file="score-set_expression_mut.csv",row.names=F)
```
