---
title: "Isogenic validation experiments"
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: inline
---

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

#list of packages to install/load
packages = c("ggplot2", "data.table", "tidyverse", "dplyr", "broom", "gridExtra")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages])
}
#load packages
invisible(lapply(packages, library, character.only=T))

#make results directory
if(!file.exists("results")){
 dir.create(file.path("results"))
}


```

## Experiment: isogenic validation of yeast-display RBD mutants and homologs

**2020-05-01**
WT SARS-CoV-2 Spike RBD was validated in triplicate. 

**2020-05-08**
WT SARS-CoV-2 Spike RBD and 6 homologs were validated in isogenic experiments. 

### Read in data table with mean bin at each concentration
```{r read_input}
dt <- read.csv(file="Table-all.csv", stringsAsFactors=F)
```

### Duplicate rows that correspond to May01 wild type triplicate experiment so we can fit models to the pooled replicates
```{r prep_triplicate_WT}
wildtype <- dt %>% 
  filter(expt=="200501") %>%
  mutate(replicate = "pooled",
         titration = "wt_pooled")

dt <- dt %>% rbind(wildtype)

print(nrow(dt))
```

### Calculate log-mean `geomean_FITC` and `FITC+` for each titration
```{r calculate_expression_values}
dt <- dt %>% 
  group_by(titration) %>%
  mutate(mean_FITCpos = mean(FITCpos),
         stderr_FITCpos = sd(FITCpos)/sqrt(length(FITCpos)),
         log_geomean_FITC = log(geomean_FITC),
         mean_logMFI_FITC = mean(log(geomean_FITC)),
         stderr_logMFI_FITC = sd(log(geomean_FITC))/sqrt(length(geomean_FITC))
         ) %>%
  ungroup()

head(dt, n=5)
```

### Use `broom` to get the results from fitting `nls` model by group
```{r fit_titrations}
nls_broom <- dt %>% 
  group_by(titration) %>%
  do(tidy(nls(mean_bin ~ a*(conc_M/(conc_M+Kd))+b,
              data=.,
              start=list(a=3,b=1,Kd=1e-10),
              lower=list(a=2,b=1,Kd=1e-15),
              upper=list(a=3,b=1.5,Kd=1e-5),
              algorithm="port"
              )
          ) 
  )

dt <- dt %>%
  merge(nls_broom %>%
          filter(term=="Kd") %>%
          select(estimate, std.error) %>%
          rename(Kd="estimate",
                 Kd_SE="std.error"), by="titration")
head(dt, n=5)
```
### Write summary table to CSV file
```{r write_output_table}
isogenic_titrations_summary <- dt %>% 
  select(expt, titration, genotype, replicate, Kd, Kd_SE, mean_FITCpos, stderr_FITCpos, mean_logMFI_FITC, stderr_logMFI_FITC) %>%
  unique()

isogenic_titrations_summary
write.csv(isogenic_titrations_summary,"./results/isogenic_titrations_summary.csv", row.names = FALSE)
```

### Now predict `mean_bin` using the models 
```{r predict_y}
conc_M = c(1:20 %o% 10^(-13:-7)) # this should only generate 120 estimates per titration (faster!)

nls_predictions <- dt %>%
  select(titration, expt, genotype, replicate) %>%
  merge(nls_broom %>%
          select(-statistic, -p.value, -std.error) %>%
          spread(term, estimate), 
        by="titration") %>%
  unique() %>%
  merge(dt %>% select(titration, Kd_SE) %>% unique(), by="titration") %>%
  merge(as.data.frame(conc_M), all=TRUE) %>%
  mutate(mean_bin = a*(conc_M/(conc_M+Kd))+b)

head(nls_predictions, n=5)
```

### Make plots for titration curves for May08 homolog experiment
```{r titrations_plot_homolog_panel, fig.width=8,fig.height=8,fig.align="center", dpi=500,dev="png",message=FALSE,error=FALSE,warning=FALSE}
annotations <- dt %>% 
  filter(expt != "200501") %>% 
  select(titration, genotype, expt, replicate, Kd, Kd_SE) %>%
  unique() %>%
  remove_rownames()

ggplot(dt %>% filter(expt != "200501"), aes(conc_M, mean_bin)) +
  geom_point() +
  geom_line(data = nls_predictions %>% filter(expt != "200501"),
            aes(conc_M, mean_bin),
            color="red") +
  scale_x_log10(lim=c(2e-14,2e-07)) +
  xlab("ACE2 (M)") +
  ylab("mean bin") +
  facet_wrap(~ genotype) +
  geom_text(
    data    = annotations,
    mapping = aes(x = 2.5e-12, 
                  y = 3.75, 
                  label = c(paste(
                    "Kd=", format(Kd, digits=2),
                    "+/-", format(Kd_SE, digits=1), "M"))),
    size=3) +
  theme_bw()

ggsave(
  "./results/homolog_titration.pdf",
  scale = 1,
  width = NA,
  height = NA,
  useDingbats=F
)
```
### Make plots just for May01 wildtype triplicate titrations
```{r wildtype_triplicate_titrations_plot, fig.width=5,fig.height=5,fig.align="center", dpi=500,dev="png",message=FALSE,error=FALSE,warning=FALSE}
annotations <- dt %>% 
  filter(expt == "200501") %>% 
  select(titration, genotype, expt, replicate, Kd, Kd_SE) %>%
  unique() %>%
  remove_rownames()

ggplot(dt %>% filter(expt == "200501"), aes(conc_M, mean_bin)) +
  geom_point() +
  geom_line(data = nls_predictions %>% filter(expt == "200501"),
            aes(conc_M, mean_bin),
            color="red") +
  scale_x_log10(lim=c(2e-14,2e-07)) +
  xlab("ACE2 (M)") +
  ylab("mean bin") +
  facet_wrap(~ replicate) + # this time facet on replicate
  geom_text(
    data    = annotations,
    mapping = aes(x = 2.5e-12, 
                  y = 3.75, 
                  label = c(paste(
                    "Kd=", format(Kd, digits=2),
                    "+/-", format(Kd_SE, digits=1), "M"))),
    size=3) +
  ggtitle("Wild Type SARS-CoV-2") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

ggsave(
  "./results/wildtype_triplicate_titration.pdf",
  scale = 1,
  width = NA,
  height = NA,
  useDingbats = F
)
```

### Plot FITC+ and geomean_FITC by genotype
We will ignore the original WT isogenic titration experiment (May01) because it is not really possible to compare across experiments due to variability with induction, etc. 
```{r expression_plots, fig.width=8,fig.height=5,fig.align="center", dpi=500,dev="png",message=FALSE,error=FALSE,warning=FALSE}
p1 <- ggplot(dt %>% filter(expt!="200501"), aes(genotype, FITCpos)) + 
  geom_boxplot() +
  geom_point() +
  scale_y_continuous(lim=c(0, 100)) +
  ylab("% FITC+") +
  facet_wrap(~ expt, scales="free_x") +
  theme_bw() +
  theme(text = element_text(size=12), 
        axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

p2 <- ggplot(dt %>% filter(expt!="200501"), aes(genotype, log_geomean_FITC)) +
  geom_boxplot() +
  geom_point() +
  ylab("mean FITC+ log-intensity") +
  facet_wrap(~ expt, scales="free_x") +
  theme_bw() +
  theme(text = element_text(size=12), 
        axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

grid.arrange(p1, p2, ncol=2, widths=c(8,8), heights=c(6))

g <- arrangeGrob(p1, p2, ncol=2, widths=c(8,8), heights=c(6))

ggsave(
  "./results/homolog_FITC_expression.pdf",
  g,
  scale = 1,
  width = NA,
  height = NA
)
```


